A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-08-19 tidyindex.qmd
first paragraph of section 2 parts moved into first paragraph of the intro

Indexes are commonly used to combine and summarize different sources of information into a single number for monitoring, communicating, and decision-making. Examples of those include the Air Quality Index, El Niño-Southern Oscillation Index, Consumer Price Index, QS University Rankings, or Human Development Index among many others.

@@

Indexes serve as critical tools to guide understanding and decisions across natural and social sciences. In environmental science, climate indexes are produces by major monitoring centres, such as the United States Drought Monitor and National Oceanic and Atmospheric Administration, to facilitate agricultural planning and early monitoring on natural disaster. In economics, indexes provide insight into market trends through combining a collection of representative items. In social sciences, indexes are useful to measure social concepts such as human development, gender equity, and university quality, which leads to country ranking and enables cross-country comparison.

While the concept of combining variables into indexes finds widespread application across various domains, indexes tend to develop separately within each field. The OECD handbook [@oecd_handbook_2008] provides a framework of steps and recommendations. for composing socio-economic composite indexes. The drought index handbook [@svoboda2016handbook], instead, provides a practical guide outlining the advantage and disadvantage of various drought indexes, as recommended by World Meteorology Organization. Review papers in the literature summarise different methodologies for constructing indexes from different variables and statistical methods, for example, @zargar2011review, @hao_drought_2015 @alahacoon_comprehensive_2022 for drought indexes. In specific domains, attention are given to analyze the index construction process, including the implications of the choices made during the index construction [@jones_vulnerability_2007] and uncertainty or sensitivity analysis [@uncertainty; @tate_social_2012].

>>>>

Indexes are commonly used to combine and summarize different sources of information into a single number for monitoring, communicating, and decision-making. They serve as critical tools across the natural and social sciences. Examples include the Air Quality Index, El Niño-Southern Oscillation Index, Consumer Price Index, QS University Rankings, and the Human Development Index. In environmental science climate indexes are produced by major monitoring centers, like the United States Drought Monitor and National Oceanic and Atmospheric Administration, to facilitate agricultural planning and early detection of natural disasters.  In economics, indexes provide insight into market trends through combining prices of a basket of goods and services. In social sciences, indexes are used to monitor human development, gender equity, or university quality. The problem is that every index is developed in its own unique way, by different researchers or organizations, and often indexes designed for the same purpose cannot easily be compared.

@@

The OECD handbook [@oecd_handbook_2008] provides a framework of steps and recommendations. for composing socio-economic composite indexes. The drought index handbook [@svoboda2016handbook], instead, provides a practical guide outlining the advantage and disadvantage of various drought indexes, as recommended by World Meteorology Organization. Review papers in the literature summarise different methodologies for constructing indexes from different variables and statistical methods, for example, @zargar2011review, @hao_drought_2015 @alahacoon_comprehensive_2022 for drought indexes. In specific domains, attention are given to analyze the index construction process, including the implications of the choices made during the index construction [@jones_vulnerability_2007] and uncertainty or sensitivity analysis [@uncertainty; @tate_social_2012].

### 2023-08-19 tidyindex.tex
first paragraph of section 2 parts moved into first paragraph of the intro

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
Indexes are useful for summarizing multivariate information into single
metrics for monitoring, communicating, and decision-making. While most
work has focused on defining new indexes for specific purposes, more
attention needs to be directed towards making it possible to understand
index behavior in different data conditions, and to determine how their
structure affects their value and variation in values. Here we discuss a
modular data pipeline recommendation to assemble indexes. It is
universally applicable to index computation and allows investigation of
index behavior as part of the development procedure. One can compute
with different the index with different parameter choices, adjust steps
in the index definition by adding, removing, and swapping them to
experiment with various index designs, calculate uncertainty measures,
and assess an index's robustness. The paper presents three examples to
illustrate the pipeline framework usage: comparison of two different
indexes designed to monitor the spatio-temporal distribution of drought
in Queensland, Australia; the effect of dimension reduction choices on
the Global Gender Gap Index (GGGI) on a country's ranking; and how to
calculate bootstrap confidence intervals for XXX index. The methods are
supported by a new R package, called \texttt{tidyindex}.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep software design\sep uncertainty\sep 
decision-making
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, interior hidden, sharp corners, frame hidden, breakable]}{\end{tcolorbox}}\fi

@@

Indexes serve as critical tools to guide understanding and decisions
across natural and social sciences. In environmental science, climate
indexes are produces by major monitoring centres, such as the United
States Drought Monitor and National Oceanic and Atmospheric
Administration, to facilitate agricultural planning and early monitoring
on natural disaster. In economics, indexes provide insight into market
trends through combining a collection of representative items. In social
sciences, indexes are useful to measure social concepts such as human
development, gender equity, and university quality, which leads to
country ranking and enables cross-country comparison.

While the concept of combining variables into indexes finds widespread
application across various domains, indexes tend to develop separately
within each field. The OECD handbook (OECD, European Union, and Joint
Research Centre - European Commission 2008) provides a framework of
steps and recommendations. for composing socio-economic composite
indexes. The drought index handbook (Svoboda, Fuchs, et al. 2016),
instead, provides a practical guide outlining the advantage and
disadvantage of various drought indexes, as recommended by World
Meteorology Organization. Review papers in the literature summarise
different methodologies for constructing indexes from different
variables and statistical methods, for example, Zargar et al. (2011),
Hao and Singh (2015) Alahacoon and Edirisinghe (2022) for drought
indexes. In specific domains, attention are given to analyze the index
construction process, including the implications of the choices made
during the index construction (Jones and Andrey 2007) and uncertainty or
sensitivity analysis (Saisana, Saltelli, and Tarantola 2005; Tate 2012).

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
Indexes are useful for summarizing multivariate information into single
metrics for monitoring, communicating, and decision-making. While most
work has focused on defining new indexes for specific purposes, more
attention needs to be directed towards making it possible to understand
index behavior in different data conditions, and to determine how their
structure affects their value and variation in values. Here we discuss a
modular data pipeline recommendation to assemble indexes. It is
universally applicable to index computation and allows investigation of
index behavior as part of the development procedure. One can compute
with different the index with different parameter choices, adjust steps
in the index definition by adding, removing, and swapping them to
experiment with various index designs, calculate uncertainty measures,
and assess an index's robustness. The paper presents three examples to
illustrate the pipeline framework usage: comparison of two different
indexes designed to monitor the spatio-temporal distribution of drought
in Queensland, Australia; the effect of dimension reduction choices on
the Global Gender Gap Index (GGGI) on a country's ranking; and how to
calculate bootstrap confidence intervals for XXX index. The methods are
supported by a new R package, called \texttt{tidyindex}.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep software design\sep uncertainty\sep 
decision-making
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, frame hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, interior hidden]}{\end{tcolorbox}}\fi

@@

The OECD handbook (OECD, European Union, and Joint Research Centre -
European Commission 2008) provides a framework of steps and
recommendations. for composing socio-economic composite indexes. The
drought index handbook (Svoboda, Fuchs, et al. 2016), instead, provides
a practical guide outlining the advantage and disadvantage of various
drought indexes, as recommended by World Meteorology Organization.
Review papers in the literature summarise different methodologies for
constructing indexes from different variables and statistical methods,
for example, Zargar et al. (2011), Hao and Singh (2015) Alahacoon and
Edirisinghe (2022) for drought indexes. In specific domains, attention
are given to analyze the index construction process, including the
implications of the choices made during the index construction (Jones
and Andrey 2007) and uncertainty or sensitivity analysis (Saisana,
Saltelli, and Tarantola 2005; Tate 2012).
